A Fredholm integral equation approach to computing Stokes eigenvalues in the plane

### 2019-03-25 doc/draft-01-stokes/03analysis.tex
Cleaning up fredholm determinant section

Let $\cJ_{1}(X)$ denote the space of trace class operators 
on $X$, where $X$ is a Hilbert space, which is a
subspace of the space of compact operators on $X$.
A compact operator $\cA$ with eigenvalues
$\lambda_{i}, i\in \mathbb{N}$ is in $\cJ_{1}(X)$ if
$\sum_{i} |\lambda_{i}| < \infty$.
If $\cA$, is a trace class operator, then  
the Fredholm determinant of the operator $\cI + \cA$
is defined by
\begin{equation}
\text{det}(\cI +\cA) = \prod_{i=1}^{\infty} (1+\lambda_{i}) \, .
\end{equation}
%It follows from standard results in complex
%analysis, that the Fredholm determinant is finite and well-defined if 
%$\text{det}(\cI+A) < \infty$ if $\sum_{i} |\lambda_{i}| < \infty$,
%i.e. if $A$ is in trace class.

The operator $-2\cD_{k} - 2\cW$ is trace class:
\begin{lemma}
  Suppose that $\Gamma$ is a $C^2$ curve.
  Then, $-2\cD_{k} - 2\cW \in \cJ_1(L_2(\Gamma)\otimes L_2(\Gamma))$
  for all $k$ 
\end{lemma}
\begin{proof}
Using Bessel function asymptotics, we note that the
kernel of $\cD_{k}$ given by $\TT_{\cdot,\cdot,\ell}\nu_{\ell}(\bx,\by)$
has a leading order singularity of
$|\bx-\by|^2 \log{|\bx-\by|^2}$ as $\bx\to \by$.
It follows from the criteria listed
in~\cite[Sec. 2]{bornemann2010numerical} that $\cD_{k}$
is a trace-class operator for all $k$.
Since $\cW$ is a rank-one perturbation independent of $k$,
and trace-class operators are a vector space, we conclude
that $-2\cD_{k} - 2\cW$ is also a trace-class operator for
all $k$.
\end{proof}

Let $f(k) = \text{det}(\cI - 2\cD_{k} - 2\cW)$.
First, note that $f(k)$ is an analytic function of $k$
for $k \in \mathbb{C} \setminus 0$, since the kernel
of $\cD_{k}$ is an analytic function of $k$ on that domain, 
and the Fredholm determinant is an analytic operator.

The zeros of the Fredholm determinant indicate when the
operator is not invertible. 
This is summarized as 
\begin{lem} \label{lem:detzeros}
  With $f(k)$ defined as above, $f(k) = 0$ if and only if
  $\cI - 2\cD_{k} -2\cW$ is not invertible.
\end{lem}
\begin{proof}
  The proof is standard; see, for example,
  \cite[p. 34]{simon2005trace}.
\end{proof}

@@

The proof presented in~\cite{zhao2015robust} carries
forward for computing the Dirichlet eigenvalues of
Stokes equation as well.
We summarize the result as
\begin{thrm}
\label{thm:mainconvfreddet}
Suppose that $\Omega$ is a simply connected
domain with an analytic boundary. Let $k_{j}^2$, $j=1,2,\ldots M$
denote all the Dirichlet eigenvalues of Stokes equation
on $\Omega$ contained in the interval $[a,b]$. 
Suppose further that all the eigenvalues have multiplicity $1$.
Let $f^{N}(k) = \text{det}(I+M^{N}_{k})$, where $M^{N}_{k}$ 
is the Nystr\"{o}m discretzation of $-2\cD_{k} - 2\cW$ with Kress
quadrature.
Then $f^{N}(k)$ has exactly $M$ zeros on the interval $[a,b]$
\note{Manas: for sufficiently large $N$?}. 
Let $\omega_{j}$, $j=1,2\ldots M$ denote the zeros of $f^{N}$.
Furthermore, there exist constants $a>0$ and $C$, 
such that $\sup_{j=1}^{M} |\omega_{j} - k_{j}| < C e^{-aN}$.
\end{thrm}

@@

\begin{remark}
In practice, using Kress-quadrature for large
problems is problematic owing to the global
nature of the quadrature rule.
%
\note{Manas: I do not understand the $O(N)$ cost per matrix entry
  described here in the fast-direct case ... does it somehow
  screw up compression?}
Coupling with fast direct solvers for computing determinants also results in $O(N^2 \log{N})$ 
determinant evaluation since evaluating each matrix entry costs $O(N)$, where 
$N$ is the number of discretization points on the boundary.
Over the last two decades, many high-order quadrature methods which are compatible with
fast algorithms like the fast multipole method (and fast direct solvers as well, 
since each matrix entry requires O(1) work) have been developed for solving
integral equations.
Extensive numerical evidence suggests that the zeros of the 
determinants of linear systems discretized using 
these quadrature methods are also high order approximations of Dirichlet eigenvalues
for the Stokes PDE--- the error is observed to be proportional to the quadrature error
for the eigenfunction $\bt^{-}$ associated with the eigenvalue.
We leave a proof of this to future work.
\end{remark}

\begin{remark}
The same analysis does not carry through for the operator
$\cI - 2 \cD_{k} - 2i\eta \cS_{k} - 2\cW$, 
since $\cS_{k}$ is not a trace class operator.
For brevity, let $\cM_{k} = -2\cD_{k} - 2i\eta \cS_{k} - 2\cW$.
The operator $\cM_{k}$ is 
in $\cJ_{2}(\bL^{2}(\Gamma))$ where
$\cJ_{2}(\bL^{2}(\Gamma))$ is the space of Hilbert-Schmidt operators
on $\bL^{2}(\Gamma)$ (the singular values of the operator are square
summable, as opposed to being summable).
Thus the Fredholm determinant of $\cI + \cM_{k}$
is not necessarily finite. 
However, as noted in~\cite{zhao2015robust}, the convergence
result~\cref{thm:mainconvfreddet}
should be true up to a logarithmic factor in the rate
of convergence, since the singular values of the operator
$\cM_{k}$ decay like $\frac{1}{n}$, and the
Fredholm determinant divergers logarithmically. 
In~\cref{subsec:convannulus}, we demonstrate this fact
numerically on the annulus, where the eigenvalues are
analytically known. 
\end{remark}

>>>>

Let $\cJ_{1}(X)$ denote the space of trace class operators 
on $X$, where $X$ is a Hilbert space, which is a
subspace of the space of compact operators on $X$.
A compact operator $\cA$ with eigenvalues
$\lambda_{i}, i\in \mathbb{N}$ is in $\cJ_{1}(X)$ if
$\sum_{i} |\lambda_{i}| < \infty$.
If $\cA$, is a trace class operator, then  
the Fredholm determinant of the operator $\cI + \cA$
is defined by
\begin{equation}
\text{det}(\cI +\cA) = \prod_{i=1}^{\infty} (1+\lambda_{i}) \, .
\end{equation}

So far, we have discussed the Fredholm theory in the space
$C(\Gamma)\times C(\Gamma)$ equipped with the bilinear form~\cref{eq:bi_form}.
However, it is more convenient to discuss the theory of Fredholm
determinants on Hilbert spaces. 
We note that both the operators $\cD_{k}$ and $\cS_{k}$ 
are also compact operators mapping $Y \to Y$
where $Y = \mathbb{L}^{2}(\Gamma) \times \mathbb{L}^{2}(\Gamma)$.
Furthermore, it is well-known that the spectrum of compact
operators with weakly singular kernels coincide on 
$C(\Gamma)\times C(\Gamma)$ and $Y$ (see~\cite{kress1989linear},
for example).
So for the rest of the section, we present the discussion of 
the relevant operators on $Y$ instead of $C(\Gamma)\times C(\Gamma)$.

%It follows from standard results in complex
%analysis, that the Fredholm determinant is finite and well-defined if 
%$\text{det}(\cI+A) < \infty$ if $\sum_{i} |\lambda_{i}| < \infty$,
%i.e. if $A$ is in trace class.

The operator $-2\cD_{k} - 2\cW$ is trace class:
\begin{lemma}
  Suppose that $\Gamma$ is a $C^2$ curve.
  Then, $-2\cD_{k} - 2\cW \in \cJ_1(Y)$
  for all $k \in \mathbb{C} \setminus \{0\}$ 
\end{lemma}
\begin{proof}
Using Bessel function asymptotics, we note that the
kernel of $\cD_{k}$ given by $\TT_{\cdot,\cdot,\ell}\nu_{\ell}(\bx,\by)$
has a leading order singularity of
$|\bx-\by|^2 \log{|\bx-\by|^2}$ as $\bx\to \by$ 
for all $k \in \mathbb{C} \setminus \{ 0\}$.
It follows from the criteria listed
in~\cite[Sec. 2]{bornemann2010numerical} that $\cD_{k}$
is a trace-class operator.
Since $\cW$ is a rank-one perturbation independent of $k$,
and trace-class operators are a vector space, we conclude
that $-2\cD_{k} - 2\cW$ is also a trace-class operator.
\end{proof}

Let $f(k) = \text{det}(\cI - 2\cD_{k} - 2\cW)$.
First, note that $f(k)$ is an analytic function of $k$
for $k \in \mathbb{C} \setminus \{0 \}$, since the kernel
of $\cD_{k}$ is an analytic function of $k$ on that domain, 
and the Fredholm determinant of an analytic operator 
is analytic on the domain of analyticity
of the operator (see~\cite{zhao2015robust}, for example).

The zeros of the Fredholm determinant indicate when the
operator is not invertible.
The following lemma summarizes this result.
\begin{lem} \label{lem:detzeros}
  With $f(k)$ defined as above, $f(k) = 0$ if and only if
  $\cI - 2\cD_{k} -2\cW$ is not invertible.
\end{lem}
\begin{proof}
  The proof is standard; see, for example,
  \cite[p. 34]{simon2005trace}.
\end{proof}

@@

The proof presented in~\cite{zhao2015robust} carries
forward for computing the Dirichlet eigenvalues of
Stokes equation as well.
The result is summarized below.
\begin{thrm}
\label{thm:mainconvfreddet}
Suppose that $\Omega$ is a simply connected
domain with an analytic boundary. Let $k_{j}^2$, $j=1,2,\ldots M$
denote all the Dirichlet eigenvalues of Stokes equation
on $\Omega$ contained in the interval $[a,b]$. 
Suppose further that all the eigenvalues have multiplicity $1$.
Let $f^{N}(k) = \text{det}(I+M^{N}_{k})$, where $M^{N}_{k}$ 
is the Nystr\"{o}m discretzation of $-2\cD_{k} - 2\cW$ with Kress
quadrature.
Then there exists $N_{0} \in \mathbb{N}$ such
that for all $N>N_{0}$, 
$f^{N}(k)$ has exactly $M$ zeros on the interval $[a,b]$.
Let $\omega_{j}$, $j=1,2\ldots M$ denote the zeros of $f^{N}$.
Furthermore, there exist constants $a>0$ and $C$, 
such that $\sup_{j=1}^{M} |\omega_{j} - k_{j}| < C e^{-aN}$.
\end{thrm}

@@

\begin{remark}
In practice, using Kress-quadrature for large
problems is problematic owing to the global
nature of the quadrature rule.
Coupling with fast direct solvers for computing determinants also results in $O(N^2 \log{N})$.
where $N$ is the number of discretization points on the boundary.
Typical fast direct solvers
query $O(N)$ matrix entries to obtain a compressed representation,
this combined with the fact that evaluating each matrix entry
costs $O(N)$ for Kress-quadratures results in the sub-optimal
complexity. 
Over the last two decades, many high-order quadrature methods which are compatible with
fast algorithms like the fast multipole method (and fast direct solvers as well, 
since each matrix entry requires O(1) work) have been developed for solving
integral equations.
Extensive numerical evidence suggests that the zeros of the 
determinants of linear systems discretized using 
these quadrature methods are also high order approximations of Dirichlet eigenvalues
for the Stokes PDE--- the error is observed to be proportional to the quadrature error
for the eigenfunction $\bt^{-}$ associated with the eigenvalue.
We leave a proof of this to future work.
\end{remark}

\begin{remark}
The same analysis does not carry through for the operator
$\cI - 2 \cD_{k} - 2i\eta \cS_{k} - 2\cW$, 
since $\cS_{k}$ is not a trace class operator.
For brevity, let $\cM_{k} = -2\cD_{k} - 2i\eta \cS_{k} - 2\cW$.
The operator $\cM_{k}$ is 
in $\cJ_{2}(Y)$ where
$\cJ_{2}(Y)$ is the space of Hilbert-Schmidt operators
on $Y$ (the singular values of the operator are square
summable, as opposed to being summable).
Thus the Fredholm determinant of $\cI + \cM_{k}$
is not necessarily finite. 
However, as noted in~\cite{zhao2015robust}, the convergence
result~\cref{thm:mainconvfreddet}
should be true up to a logarithmic factor in the rate
of convergence, since the singular values of the operator
$\cM_{k}$ decay like $\frac{1}{n}$, and the
Fredholm determinant divergers logarithmically. 
In~\cref{subsec:convannulus}, we demonstrate this fact
numerically on the annulus, where the eigenvalues are
analytically known. 
\end{remark}
